$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-07-28 qdpm_paper.tex
new struc factor cut plot
Worked on text

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\secref}[1]{Sec. \ref{#1}}
\newcommand{\tabref}[1]{Tab. \ref{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t!]
    \centering
    \includegraphics[width=1.0\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The gauge symmetry allows one to construct a gauge invariant string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv -\frac{c_P}{4} \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) 
\end{equation}
and the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma+ \tilde{\Phi}_\Gamma$. (Alternately can think of the source of $\Phi_\Gamma$ to be static charges of charge $+c_0$ ($-c_0$) sitting at the verticies of the $A$($B$) sub-lattice.).

@@

 %%%%%%%%%%%%%%%%%
%\begin{figure}[tpb]
%\centering
%\includegraphics[width=0.8\linewidth]{pentamer_moves.pdf}
%\caption{a) Horizontal creation an annihilation, b) horizontal move \textcolor{red}{Owen:fix this move}, c) diagonal moves \textcolor{red}{Owen: Remove (d) as it is related to (c) by rotational symmetry}}
%\label{fig:pentamer_moves}
%\end{figure}
%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}
%    \begin{tabular}{| c | c | c | c | c | c |}
%  	\hline
%    $c_1$& $\bar{c}_1$ & $c_2$& $\bar{c}_2$ & $c_2$& $\bar{c}_2$ \\
%	\hline
%    $\pentmvcreate$& $\pentmvanal$ & $\penmvhrzleft$& $\penmvhrzright$ & $\penmvdiagup$& $\penmvdiagdown$ \\
%	\hline
%  \end{tabular}
%  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
%  \label{tab:defects}
%\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
     & $c$ & $\bar{c}$ \\
	\hline
    1 & $\pentmvcreate$ &  $\pentmvanal$  \\
	\hline
    2 &$\penmvhrzleft$&$\penmvhrzright$ \\
	\hline
    3 &$\penmvdiagup$& $\penmvdiagdown$\\
	\hline
  \end{tabular}
  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}

@@

To determine the nature of the ground state at the RK point of the QDPM, we have performed Monte Carlo calculations of $\ket{\Psi_{\rm{RK}}}$, using local updates corresponding to the dimer resonances shown in \figref{fig:pentamer_moves}.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the $U(1)$ topological winding number $\Phi_X$ histogram. This is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors. We confirm that $\Phi_{X,Y}$ are conserved modulo $3$, leading to $9$-fold topological degeneracy. Notice that $\Phi_X = \pm L/2$ correspond to ordered configurations, and therefore the probability of $\Phi_{XY}$ reaching these extrema vanishes in the RK state.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ winding number $\Phi_X$ as measured in the QDPM on a
    $32\times 32$ lattice. Different colors correspond to calculations initialized to different winding sectors. Note that $\Phi_X$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the possibility of symmetry breaking order in the RK ground state of the QDPM, we have computed the dimer-dimer correlation function:
\begin{equation}
    C_{\mathrm{d}} = \langle n_0 n_r \rangle - \langle n_0 \rangle   \langle n_r \rangle   
    .
\end{equation}
\figref{fig:dimer_gry_scale_qdpm} shows $C_d$ in real space for and $L=16$ lattice; we see no evidence of symmetry breaking in systems up to $L=64$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel dimers is plotted as a function of separation, and displays an exponential decay with correlation length $\chi \approx ?????$. Thus we find that the RK ground state of the QDPM is a symmetric dimer-pentamer liquid with exponentially decaying correlations.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{Gray scale of the dimer (left) and $Z_3$ vison (right) correlation functions
            The color blue indicates the dimer (plaquete) from which the correlation is measured. The 
            color red indicate correlations that are off the gray scale used here .}
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_parallel.pdf}
    \caption{Correlation function between parallel dimers on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice. }
    \label{fig:spatial_dmr_cor}
\end{figure}
%    The lines represent least-squares fits to exponential decay. }
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_perp.pdf}
    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_z3_vis.pdf}
    \caption{ $Z_3$ vison correlation function for $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

   \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{qdpm_dmr_vis_HM_16x16.pdf}
        \caption{QDPM dimer structure factor $\gamma=0.5$.}
        \label{fig:spacial_dmr_z3vis_cor}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{vis_dmr_qdpm_struc_fac_16x16.pdf}
        \caption{The QDPM dimer structure factor $S_d(k)$ and $Z_3$ vison structure factor $S_v(k)$
        shown along the path in $k$ space connecting the $(0,0)$, to $(\pi,0)$, $(\pi,\pi)$, $(0,0)$
        points for $\gamma=0.5$.}
        \label{fig:vis_dmr_cut}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{Monomer correlations in the QDPM.}
        \label{fig:vis_dmr_cut}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\secref}[1]{Sec. \ref{#1}}
\newcommand{\tabref}[1]{Tab. \ref{#1}}
\newcommand{\note}[1]{\textcolor{red}{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.75\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The gauge symmetry allows one to construct a gauge invariant string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv - \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell
\end{equation}
where $n_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge, and the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma+ \tilde{\Phi}_\Gamma$. (Alternately can think of the source of $\Phi_\Gamma$ to be static charges of charge $+c_0$ ($-c_0$) sitting at the verticies of the $A$($B$) sub-lattice.).

@@

\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
     & $c$ & $\bar{c}$ \\
	\hline
    $i$ & $\pentmvcreate$ &  $\pentmvanal$  \\
	\hline
    $ii$ &$\penmvhrzleft$&$\penmvhrzright$ \\
	\hline
    $iii$ &$\penmvdiagup$& $\penmvdiagdown$\\
	\hline
  \end{tabular}
  \caption{The pairs of local dimer \& pentamer configurations $c\leftrightarrow\bar{c}$ generate minimal local dynamics that create/annihilate pairs ($i$) of pentamers and translate single pentamers ($ii$ \& $iii$). Note that pentamers always stay on the same sublattice which is associates with their $U(1)$ charge. }
  \label{tab:pentamerMoves}
\end{table}

@@

To determine the nature of the ground state at the RK point of the QDPM, we have used Monte Carlo sampling of $\ket{\Psi_{\rm{RK}}}$, using local updates corresponding to the dimer resonances shown in \tabref{tab:pentamerMoves} to compute ground state expectation values. We studied $L\times L$ lattices with periodic boundary conditions on systems up to $L=64$.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding number $\Phi_X$ . This is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors; we confirm that $\Phi_{X/Y}$ is only conserved modulo $3$, rather than strictly conserved as it is in the QDM. We find that $P(\Phi_{X/Y})$ has a roughly gaussian envelope with a width that scales with system size--this is expected as $\Phi_{X/Y} = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled by $\Phi_{X/Y} \mod 3$, or equivalently $(W_X^{\rm{QDPM}},W_Y^{\rm{QDPM}})$, in contrast with the extensive degeneracy of the QDM. For the remainder of the paper we present results in the $(0,0)$, topological sector, but we see no dependence the topological sector for any of the results presented here~\note{Owen: which calculations/system sizes have you checked different topological sectors?}.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ winding number $\Phi_X$ as measured in the QDPM on a
    $L=32$ lattice. Different colors correspond to calculations initialized to different winding sectors. Note that $\Phi_X$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.
    \note{Owen: change the y-axis label to $P(\Phi_x)$ and the x-axis to $\Phi_x$. Is this still an $L=32$ lattice? Do you have $L=64$? Perhaps put both on the same plot? (with no vertical bars)}}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Here we discuss the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_gry_scale_qdpm} we plot the bare dimer-dimer correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=16$ lattice; here the origin $\bm{0}$ represents the blue link, and $\bm{r}$ varies over all other links of the lattice. We see no evidence of symmetry breaking in the dimer correlations on lattices up $L=64$.

  \begin{equation}
    C_{\mathrm{d}} = \langle n_0 n_r \rangle - \langle n_0 \rangle   \langle n_r \rangle   
    .
\end{equation}
 In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel dimers is plotted as a function of separation, and displays an exponential decay with correlation length $\chi \approx ?????$. Thus we find that the RK ground state of the QDPM is a symmetric dimer-pentamer liquid with exponentially decaying correlations.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{({\it left}) Gray scale of the dimer-dimer correlation function $\langle n_0 n_{\bm{r}}\rangle$ on an $L=16$ lattice. The origin corresponds to the blue link, and $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off the given scale.~\note{Owen: Is this the bare or connected correlation function? What do these plots look like on a log scale?} ({\it right}) $Z_3$ magnetic string correlation function. }
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_parallel.pdf}
    \caption{Correlation function between parallel dimers on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice. \note{Owen: Which direction is this? Do you have data for the perpendicular direction?}}
    \label{fig:spatial_dmr_cor}
\end{figure}
%    The lines represent least-squares fits to exponential decay. }
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_perp.pdf}
    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_cors_z3_vis.pdf}
    \caption{ $Z_3$ vison correlation function for $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

   \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{qdpm_dmr_vis_HM_16x16.pdf}
        \caption{QDPM dimer structure factor $\gamma=0.5$.}
        \label{fig:spacial_dmr_z3vis_cor}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{vis_dmr_qdpm_struc_fac_32x32.pdf}
        \caption{The QDPM dimer structure factor $S_d(k)$ and $Z_3$ vison structure factor $S_v(k)$
        shown along the path in $k$ space connecting the $(0,0)$, to $(\pi,0)$, $(\pi,\pi)$, $(0,0)$
        points for $\gamma=0.5$.}
        \label{fig:vis_dmr_cut}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{Monomer correlations in the QDPM.}
        \label{fig:vis_dmr_cut}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
